---
title: "Perry Map Products"
subtitle: "UGA GrandFarm -- Perry, Georgia"
author: "Madelyn Willis"
date: today
format:
  pdf:
    documentclass: article
    papersize: letter
    geometry:
      - top=0.6in
      - bottom=0.6in
      - left=0.6in
      - right=0.6in
    fig-pos: "H"
    keep-tex: false
    colorlinks: true
execute:
  echo: false
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    dev: "ragg_png"
    dpi: 300
---

```{r setup}
#| include: false

## ---------------------------------------------------------------------------
## Packages
## ---------------------------------------------------------------------------
## install.packages(c("terra","sf","tmap","ragg","grid","RColorBrewer"))
library(terra)
library(sf)
library(tmap)
library(grid)

sf::sf_use_s2(FALSE)
terra::terraOptions(progress = 0)

## tmap v3 and v4 have different APIs. Everything below routes through the
## helpers in the next chunk so the document renders under either version.
TMAP4 <- utils::packageVersion("tmap") >= "4.0.0"
message("tmap version detected: ", as.character(utils::packageVersion("tmap")),
        if (TMAP4) "  (v4 API)" else "  (v3 API)")

tmap_mode("plot")

## In v4, components rescale themselves per panel. In a 3-panel arrangement the
## rescaling can collapse a legend to zero/NA width, which throws
##   Error in if (tail(xright, 1) > W) : missing value where TRUE/FALSE needed
## Turning autoscale off keeps component sizing deterministic.
if (TMAP4) tmap_options(component.autoscale = FALSE)
```

```{r smoke-test}
#| include: false

## Renders one map with default settings. If THIS fails the problem is the data
## or the tmap install, not the styling below. Set include: true to inspect.
.smoke <- function() {
  r <- terra::rast(file.path(dir_krige, "chem_Ca_OK_lbsac.tif"))
  print(tm_shape(r) + tm_raster())
  invisible(TRUE)
}
```

```{r paths}
#| include: false

## ---------------------------------------------------------------------------
## PATHS -- edit these three if the projects move
## ---------------------------------------------------------------------------
dir_krige  <- "C:/git/spatial_SCP_taxonomy/outputs/perry_ordinary_kriging_no_covars/clipped_and_lbsac"
dir_series <- "C:/git/coastal_plain_taxonomy/outputs/spatial"
path_border <- "C:/Users/Madelyn Willis/OneDrive/Documents/perry_git/perry/Mapping/Data/Border/border_17N.shp"

## WHERE THE PDFs GO
dir_out <- "C:/git/spatial_SCP_taxonomy/outputs/maps"
dir.create(dir_out, recursive = TRUE, showWarnings = FALSE)

stopifnot(dir.exists(dir_krige), dir.exists(dir_series), file.exists(path_border))

## ---------------------------------------------------------------------------
## Site border -- NAD83 / UTM 17N (EPSG:26917), same CRS as the kriged rasters
## ---------------------------------------------------------------------------
border <- sf::st_read(path_border, quiet = TRUE)
border <- sf::st_transform(border, 26917)
border <- sf::st_zm(border, drop = TRUE)          # drop Z/M left over from the KML export
TARGET_CRS <- 26917
```

```{r helpers}
#| include: false

## ---------------------------------------------------------------------------
## Version-safe tmap wrappers
## ---------------------------------------------------------------------------

## Palette names differ between versions: v4 uses "brewer.yl_gn_bu",
## v3 wants the classic RColorBrewer name "YlGnBu".
pal_v3 <- c(
  "brewer.yl_gn_bu" = "YlGnBu", "brewer.pu_bu_gn" = "PuBuGn",
  "brewer.yl_or_br" = "YlOrBr", "brewer.yl_gn"    = "YlGn",
  "brewer.rd_yl_bu" = "RdYlBu", "brewer.bu_pu"    = "BuPu",
  "brewer.bu_gn"    = "BuGn",   "brewer.or_rd"    = "OrRd",
  "viridis"         = "viridis"
)
resolve_pal <- function(pal) {
  if (TMAP4) return(pal)
  if (pal %in% names(pal_v3)) unname(pal_v3[pal]) else pal
}

## Continuous raster layer with a readable legend.
##
## NOTE: legend titles must be SINGLE-LINE. tmap measures a legend title as one
## string; an embedded "\n" makes the measured width NA, which surfaces later as
##   Error in if (tail(xright, 1) > W) : missing value where TRUE/FALSE needed
## Put units in parentheses on the same line instead.
tm_raster_safe <- function(pal, title, n_breaks = 6) {
  pal   <- resolve_pal(pal)
  title <- gsub("[\r\n]+", " ", title)
  if (TMAP4) {
    tm_raster(
      col.scale  = tm_scale_intervals(n = n_breaks, style = "quantile", values = pal),
      col.legend = tm_legend(title = title, reverse = TRUE,
                             position = tm_pos_out("right", "center"))
    )
  } else {
    tm_raster(
      style   = "quantile",
      n       = n_breaks,
      palette = pal,
      title   = title,
      legend.reverse = TRUE
    )
  }
}

## Categorical polygon fill with an explicit named palette
tm_polys_safe <- function(field, pal, title, legend_ncol = 1,
                          border_col = "grey25", border_lwd = 0.35) {
  title <- gsub("[\r\n]+", " ", title)
  if (TMAP4) {
    tm_polygons(
      fill       = field,
      ## no value.na / label.na -- every polygon is recoded to a real class,
      ## so there is no "No data" entry in the legend
      fill.scale = tm_scale_categorical(values = pal),
      fill.legend = tm_legend(title = title, ncol = legend_ncol,
                              position = tm_pos_out("right", "center")),
      col        = border_col,
      lwd        = border_lwd
    )
  } else {
    tm_polygons(
      col          = field,
      palette      = pal,
      title        = title,
      border.col   = border_col,
      lwd          = border_lwd,
      colorNA      = "grey85",
      textNA       = "No data",
      legend.is.portrait = TRUE
    )
  }
}

## Scale bar -- renamed between v3 and v4.
##
## IMPORTANT: do NOT hardcode `breaks`. If the requested breaks are wider than
## the panel, tmap warns "Not all scale bar breaks could be plotted" and then
## immediately fails with
##   Error in if (tail(xright, 1) > W) : missing value where TRUE/FALSE needed
## because the unplaced break leaves xright = NA. Letting tmap choose breaks
## from the actual panel width avoids this at any figure size.
##
## If you want fixed breaks back, pass e.g. breaks = c(0, 250, 500) AND raise
## the chunk's fig-width, then confirm no warning appears.
## Placed OUTSIDE the map frame, underneath the map shape, via tm_pos_out() --
## an inside position overlaps the southern edge of the site polygon.
tm_scalebar_safe <- function(breaks = NULL, position = NULL, text.size = 0.6) {
  if (TMAP4) {
    if (is.null(position)) position <- tm_pos_out("center", "bottom")
    if (is.null(breaks)) {
      tm_scalebar(position = position, text.size = text.size)
    } else {
      tm_scalebar(breaks = breaks, position = position, text.size = text.size)
    }
  } else {
    if (is.null(position)) position <- c("left", "bottom")
    if (is.null(breaks)) {
      tm_scale_bar(position = position, text.size = text.size)
    } else {
      tm_scale_bar(breaks = breaks, position = position, text.size = text.size)
    }
  }
}

## Compass rose
tm_compass_safe <- function(position = c("right", "top"), size = 2.2) {
  if (TMAP4) {
    tm_compass(type = "rose", size = size, position = position, text.size = 0.55)
  } else {
    tm_compass(type = "rose", size = size, position = position,
               text.size = 0.55, show.labels = 1)
  }
}

## ---------------------------------------------------------------------------
## Label placement.
##
## tm_text() on a polygon layer uses the polygon CENTROID, which for concave or
## crescent-shaped map units can fall OUTSIDE the polygon it belongs to.
## st_point_on_surface() is guaranteed to return a point inside the geometry,
## so labels always sit in the unit they describe.
##
## Rows with a blank / "Unknown" label are dropped here rather than drawn.
## ---------------------------------------------------------------------------
label_points <- function(x, field) {
  v <- trimws(as.character(x[[field]]))
  keep <- !is.na(v) & v != "" & !toupper(v) %in% c("UNKNOWN", "NA", "<NA>")
  x <- x[keep, ]
  if (nrow(x) == 0) return(x)
  suppressWarnings(sf::st_point_on_surface(x))
}

## ---------------------------------------------------------------------------
## Dissolve polygons by an attribute, so shared internal edges disappear.
##
## SSURGO_matchup is an INTERSECTION of the SSURGO units with our soil series
## polygons, so its outlines are a mixture of both line networks. Dissolving on
## MUKEY collapses the intersection slivers and recovers the true SSURGO unit
## boundaries, which can then be drawn on their own.
## ---------------------------------------------------------------------------
dissolve_by <- function(x, field) {
  keys  <- as.character(x[[field]])
  keys[is.na(keys) | keys == ""] <- "__blank__"
  uk    <- unique(keys)
  parts <- lapply(uk, function(k) sf::st_union(x[keys == k, ]))
  sf::st_sf(
    stats::setNames(data.frame(uk, stringsAsFactors = FALSE), field),
    geometry = do.call(c, parts),
    crs      = sf::st_crs(x)
  )
}

## Manual legend entry for line styles -- "line" in v3, "lines" in v4
tm_line_legend_safe <- function(labels, col, lty, lwd, title = "") {
  if (TMAP4) {
    tm_add_legend(type = "lines", labels = labels, col = col,
                  lty = lty, lwd = lwd, title = title,
                  position = tm_pos_out("right", "center"))
  } else {
    tm_add_legend(type = "line", labels = labels, col = col,
                  lty = lty, lwd = lwd, title = title)
  }
}

## Polygon labels (map unit symbols)
tm_text_safe <- function(field, size = 0.4) {
  if (TMAP4) {
    tm_text(text = field, size = size, col = "grey10")
  } else {
    tm_text(text = field, size = size, col = "grey10", shadow = TRUE)
  }
}

## Shared title + layout used by every panel.
## v4 moved the panel title out of tm_layout() into its own tm_title() layer.
panel_layout <- function(title) {
  title <- gsub("[\r\n]+", " ", title)
  if (TMAP4) {
    tm_title(title, size = 0.9, fontface = "bold") +
      tm_layout(
        legend.title.size = 0.75,
        legend.text.size  = 0.58,
        frame         = TRUE,
        frame.lwd     = 0.8,
        inner.margins = c(0.03, 0.03, 0.06, 0.03),
        outer.margins = c(0.01, 0.01, 0.01, 0.01),
        bg.color      = "white"
      )
  } else {
    tm_layout(
      main.title          = title,
      main.title.size     = 0.9,
      main.title.fontface = "bold",
      main.title.position = "left",
      legend.outside          = TRUE,
      legend.outside.position = "right",
      legend.outside.size     = 0.26,
      legend.title.size       = 0.75,
      legend.text.size        = 0.58,
      frame         = TRUE,
      frame.lwd     = 0.8,
      inner.margins = c(0.03, 0.03, 0.03, 0.03),
      outer.margins = c(0.01, 0.01, 0.01, 0.01),
      bg.color      = "white"
    )
  }
}

## ---------------------------------------------------------------------------
## One continuous raster -> one finished map panel
## ---------------------------------------------------------------------------
make_raster_map <- function(file, title, legend_title, pal = "viridis",
                            band = 1, n_breaks = 6, mult = 1) {
  r <- terra::rast(file.path(dir_krige, file))
  if (terra::nlyr(r) > 1) r <- r[[band]]
  if (mult != 1) r <- r * mult
  if (!terra::same.crs(r, paste0("EPSG:", TARGET_CRS))) {
    r <- terra::project(r, paste0("EPSG:", TARGET_CRS))
  }
  ## Layer name feeds tmap's default legend title -- keep it a plain,
  ## single-line, syntactically valid name.
  names(r) <- make.names(gsub("[\r\n]+", " ", legend_title))

  tm_shape(r, bbox = sf::st_bbox(border)) +
    tm_raster_safe(pal = pal, title = legend_title, n_breaks = n_breaks) +
    tm_shape(border) +
    tm_borders(col = "black", lwd = 1.6) +
    tm_compass_safe() +
    tm_scalebar_safe() +
    panel_layout(title)
}

## ---------------------------------------------------------------------------
## Print a list of maps three to a page.
##
## Each map is printed on its own -- pagination is handled by LaTeX via an
## explicit \newpage after every third figure, NOT by tmap_arrange(). This is
## deliberate: tmap_arrange() re-runs component layout across panels and is the
## suspected source of
##   Error in if (tail(xright, 1) > W) : missing value where TRUE/FALSE needed
## Chunk fig-height controls the per-map height (2.9in x 3 fits one page).
## ---------------------------------------------------------------------------
print_three_per_page <- function(maps) {
  for (i in seq_along(maps)) {
    print(maps[[i]])
    cat("\n\n")
    if (i %% 3 == 0 && i < length(maps)) cat("\\newpage\n\n")
  }
}

## ---------------------------------------------------------------------------
## Write a PDF straight into dir_out, 3 maps per page.
## This does NOT depend on rendering the document -- you can just run the
## chunks (Run > Run All) and the PDFs appear in outputs/maps.
## ---------------------------------------------------------------------------
save_three_per_page <- function(maps, file, width = 7, height = 9.5) {
  path <- file.path(dir_out, file)
  grDevices::pdf(path, width = width, height = height, onefile = TRUE)
  on.exit(grDevices::dev.off(), add = TRUE)
  groups <- split(maps, ceiling(seq_along(maps) / 3))
  for (g in groups) print(tmap_arrange(g, nrow = 3, ncol = 1))
  message("wrote ", path)
  invisible(path)
}

## One map, one PDF
save_one <- function(map, file, width = 7.2, height = 8.4) {
  path <- file.path(dir_out, file)
  grDevices::pdf(path, width = width, height = height)
  on.exit(grDevices::dev.off(), add = TRUE)
  print(map)
  message("wrote ", path)
  invisible(path)
}
```

```{r layer-registry}
#| include: false

## ---------------------------------------------------------------------------
## LAYER REGISTRY -- one row per map. Add, remove or reorder freely.
##   file         : filename inside dir_krige
##   title        : panel heading
##   legend_title : legend heading (put units here)
##   pal          : any tmap / viridis / RColorBrewer palette name
##   band         : layer index for multi-band .tifs
## ---------------------------------------------------------------------------

## ---------------------------------------------------------------------------
## UNIT CONVERSION -- READ BEFORE CHANGING
##
## The files in clipped_and_lbsac are NAMED "_lbsac" but hold ppm. In
## scp_spatial_prediction_perry_06_30_2026.qmd two blocks write to that folder:
## the first multiplies by 2 and appends "_lbsac"; the second appends "_lbsac"
## but writes the clipped raster unconverted, deleting the first block's file
## on the way. The second block ran last, so the values are ppm.
##
## Measured means bear this out (Zn 2.0, Mn 14.5, P 17.4, Ca 376) -- normal
## Mehlich-1 ppm, far too low for lbs/ac.
##
## So we convert on read: lbs/ac = ppm x 2, the standard acre-furrow-slice
## factor (6-2/3 in depth ~ 2,000,000 lb soil/acre) that UGA AESL uses.
##
## *** If you ever fix the source pipeline so the .tifs really are lbs/ac,
## *** set CONVERT_PPM_TO_LBSAC <- FALSE or the maps will double-count.
##
## pH and LBC are NOT converted -- pH is a log scale and LBC is a buffer
## index. The source script excludes both as well.
## ---------------------------------------------------------------------------
CONVERT_PPM_TO_LBSAC <- TRUE
PPM2 <- if (CONVERT_PPM_TO_LBSAC) 2 else 1
chem_unit <- if (CONVERT_PPM_TO_LBSAC) "lbs/ac" else "ppm"

## legend_title must stay single-line -- see tm_raster_safe().
##
## PAGE BREAKS FOLLOW THIS ORDER: maps are printed in list order and a new page
## starts after every third one. So entries 1-3 share page 1, 4-6 page 2, etc.
## To regroup a page, just move entries within this list.
##
##   Page 1: Ca, pH, LBC   (the liming set -- read together)
##   Page 2: K, Mg, Mn
##   Page 3: P, Zn
chem_layers <- list(
  ## -- page 1 --------------------------------------------------------------
  list(file = "chem_Ca_OK_lbsac.tif",  title = "Calcium (Ca)", mult = PPM2,
       legend_title = paste0("Ca (", chem_unit, ")"), pal = "brewer.yl_gn_bu"),
  list(file = "chem_pH.2_OK.tif",      title = "Soil pH",
       legend_title = "pH",            pal = "brewer.rd_yl_bu"),
  list(file = "chem_LBC_1_OK.tif",     title = "Lime Buffer Capacity",
       legend_title = "LBC",           pal = "brewer.pu_bu_gn"),
  ## -- page 2 --------------------------------------------------------------
  list(file = "chem_K_OK_lbsac.tif",   title = "Potassium (K)", mult = PPM2,
       legend_title = paste0("K (", chem_unit, ")"),  pal = "brewer.yl_gn_bu"),
  list(file = "chem_Mg_OK_lbsac.tif",  title = "Magnesium (Mg)", mult = PPM2,
       legend_title = paste0("Mg (", chem_unit, ")"), pal = "brewer.yl_gn_bu"),
  list(file = "chem_Mn_OK_lbsac.tif",  title = "Manganese (Mn)", mult = PPM2,
       legend_title = paste0("Mn (", chem_unit, ")"), pal = "brewer.yl_or_br"),
  ## -- page 3 --------------------------------------------------------------
  list(file = "chem_P_OK_lbsac.tif",   title = "Phosphorus (P)", mult = PPM2,
       legend_title = paste0("P (", chem_unit, ")"),  pal = "brewer.yl_gn"),
  list(file = "chem_Zn_OK_lbsac.tif",  title = "Zinc (Zn)", mult = PPM2,
       legend_title = paste0("Zn (", chem_unit, ")"), pal = "brewer.bu_pu")
)

## Texture: the normalised 3-band stack sums to 100% across sand/silt/clay.
## To use the raw lab surfaces instead, swap in texture_lab_sand_OK.tif,
## texture_lab_silt_OK.tif, texture_lab_clay_OK.tif with band = 1.
texture_layers <- list(
  list(file = "texture_OK_normalized_0_100_sum100.tif", band = 1,
       title = "Sand", legend_title = "Sand (%)", pal = "brewer.yl_or_br"),
  list(file = "texture_OK_normalized_0_100_sum100.tif", band = 2,
       title = "Silt", legend_title = "Silt (%)", pal = "brewer.bu_gn"),
  list(file = "texture_OK_normalized_0_100_sum100.tif", band = 3,
       title = "Clay", legend_title = "Clay (%)", pal = "brewer.or_rd")
)

build_maps <- function(spec) {
  lapply(spec, function(s) {
    make_raster_map(
      file         = s$file,
      title        = s$title,
      legend_title = s$legend_title,
      pal          = s$pal,
      band         = if (is.null(s$band)) 1 else s$band,
      mult         = if (is.null(s$mult)) 1 else s$mult
    )
  })
}
```

```{r series-palette}
#| include: false

## ---------------------------------------------------------------------------
## Soil series palette, sampled from the ArcGIS Pro legend so the PDF matches
## the project's existing figures. Unlisted series fall back to a
## colourblind-safe ramp (see series_palette_for()).
## ---------------------------------------------------------------------------
arcgis_series_cols <- c(
  "Disturbed"   = "#886868",
  "Benevolence" = "#EEDABA",
  "Blanton"     = "#B68870",
  "Dothan"      = "#E6E600",
  "Faceville"   = "#FF5500",
  "Greenville"  = "#E64C00",
  "Johns"       = "#ED7932",
  "Lakeland"    = "#FFEBAF",
  "Leefield"    = "#F5B05C",
  "Lucy"        = "#F57A7A",
  "Marvyn"      = "#BA7049",
  "Norfolk"     = "#FFAA00",
  "Orangeburg"  = "#E69800",
  "Red Bay"     = "#A83800",
  "Thursa"      = "#F79965",
  "Troup"       = "#A87000",
  ## ArcGIS gave Wagram the same swatch as Lakeland; nudged so the two are
  ## distinguishable in print. Set to "#FFEBAF" to match ArcGIS exactly.
  "Wagram"      = "#E8D08A",
  ## HTM map unit -- the grey that ArcGIS used for the unnamed polygons
  "HTM"         = "#D9D9D9"
)

## Okabe-Ito extended -- colourblind-safe fallback for any unmatched class
cb_safe <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2",
             "#D55E00", "#CC79A7", "#999999", "#8C510A", "#01665E",
             "#762A83", "#1B7837", "#B2182B", "#2166AC", "#F1A340",
             "#998EC3", "#40004B")

series_palette_for <- function(levels_vec) {
  levels_vec <- levels_vec[!is.na(levels_vec)]
  pal <- setNames(rep(NA_character_, length(levels_vec)), levels_vec)
  hit <- levels_vec %in% names(arcgis_series_cols)
  pal[hit] <- arcgis_series_cols[levels_vec[hit]]
  n_miss <- sum(!hit)
  if (n_miss > 0) {
    pal[!hit] <- rep(cb_safe, length.out = n_miss)[seq_len(n_miss)]
    message("Series with no ArcGIS colour, using colourblind-safe fallback: ",
            paste(levels_vec[!hit], collapse = ", "))
  }
  pal
}

## Look for a colour column carried on the shapefile itself before falling
## back to the ArcGIS-sampled lookup.
colour_field_in <- function(x) {
  cand <- grep("^(color|colour|hex|rgb|symbol.*col)", names(x),
               ignore.case = TRUE, value = TRUE)
  if (!length(cand)) return(NULL)
  for (f in cand) {
    v <- as.character(x[[f]])
    if (all(grepl("^#?[0-9A-Fa-f]{6}$", v[!is.na(v)]))) return(f)
  }
  NULL
}
```

\newpage

# Soil Chemistry

```{r chem-maps}
#| fig-width: 7.0
#| fig-height: 2.9
#| results: asis

chem_maps <- build_maps(chem_layers)
save_three_per_page(chem_maps, "Perry_chemistry_maps.pdf")   # -> outputs/maps
print_three_per_page(chem_maps)
```

\newpage

# Particle Size Distribution

Sand, silt and clay from the normalised kriged surfaces (each cell sums to 100%).

```{r texture-maps}
#| fig-width: 7.0
#| fig-height: 2.9
#| results: asis

texture_maps <- build_maps(texture_layers)
save_three_per_page(texture_maps, "Perry_texture_maps.pdf")  # -> outputs/maps
print_three_per_page(texture_maps)
```

\newpage

# Soil Series Map

```{r series-map}
#| fig-width: 7.2
#| fig-height: 8.4

series <- sf::st_read(file.path(dir_series, "Soil_Series_Map.shp"), quiet = TRUE)
series <- sf::st_zm(series, drop = TRUE)
series <- sf::st_transform(series, TARGET_CRS)          # from GA West State Plane (ft)
series <- sf::st_make_valid(series)

## Field that carries the series name in this shapefile
SERIES_FIELD <- "Series_pri"
series[[SERIES_FIELD]] <- trimws(as.character(series[[SERIES_FIELD]]))

## Polygons with no series name are the HTM map unit. Label them HTM rather
## than letting tmap render them as a grey "No data" class.
blank_series <- is.na(series[[SERIES_FIELD]]) | series[[SERIES_FIELD]] == ""
series[[SERIES_FIELD]][blank_series] <- "HTM"

## HTM is a map unit, not a soil series, so sort it to the bottom of the legend
lv <- sort(unique(series[[SERIES_FIELD]]))
lv <- c(setdiff(lv, "HTM"), intersect(lv, "HTM"))
series[[SERIES_FIELD]] <- factor(series[[SERIES_FIELD]], levels = lv)

## Label points: drops blank/"Unknown" symbols and forces each remaining label
## inside its own polygon.
series_lab <- label_points(series, "musym")

cf <- colour_field_in(series)
if (!is.null(cf)) {
  lv  <- levels(series[[SERIES_FIELD]])
  pal <- vapply(lv, function(l) {
    v <- as.character(series[[cf]][series[[SERIES_FIELD]] == l])[1]
    if (substr(v, 1, 1) == "#") v else paste0("#", v)
  }, character(1))
  message("Using colours from shapefile field: ", cf)
} else {
  pal <- series_palette_for(levels(series[[SERIES_FIELD]]))
}

series_map <- tm_shape(series, bbox = sf::st_bbox(border)) +
  tm_polys_safe(SERIES_FIELD, pal, "Soil series", legend_ncol = 1) +
  tm_shape(series_lab) +
  tm_text_safe("musym", size = 0.38) +
  tm_shape(border) +
  tm_borders(col = "black", lwd = 1.8) +
  tm_compass_safe(size = 2.6) +
  tm_scalebar_safe() +
  panel_layout("Mapped Soil Series -- UGA GrandFarm, Perry GA")

save_one(series_map, "Perry_soil_series.pdf")                # -> outputs/maps
series_map
```

\newpage

# SSURGO Matchup

Polygons are coloured by whether the field-mapped soil series agrees with the
SSURGO map unit at that location: green for agreement, orange for disagreement
(disagreement includes polygons with no SSURGO overlap). Labels are the map
unit symbols of the soil series actually present at Perry, not SSURGO's.
Dashed grey lines are SSURGO map unit boundaries; solid black lines are our
mapped soil series boundaries.

```{r ssurgo-map}
#| fig-width: 7.2
#| fig-height: 8.4

ssurgo <- sf::st_read(file.path(dir_series, "SSURGO_matchup.shp"), quiet = TRUE)
ssurgo <- sf::st_zm(ssurgo, drop = TRUE)
ssurgo <- sf::st_transform(ssurgo, TARGET_CRS)          # from Web Mercator
ssurgo <- sf::st_make_valid(ssurgo)

## Collapse to two classes: anything that is not an explicit AGREE -- including
## NO_OVERLAP and missing values -- counts as Disagree.
MATCH_FIELD <- "match2"
raw_match <- toupper(trimws(as.character(ssurgo$match_stat)))
ssurgo[[MATCH_FIELD]] <- factor(
  ifelse(!is.na(raw_match) & raw_match == "AGREE", "Agree", "Disagree"),
  levels = c("Agree", "Disagree")
)

match_pal <- c(
  "Agree"    = "#1B7837",   # green
  "Disagree" = "#E8710A"    # orange
)

## ---------------------------------------------------------------------------
## Labels: reuse the EXACT label layer drawn on the soil series map.
##
## `series_lab` is already the Perry map unit symbols, one point per polygon,
## each placed inside its own polygon. Drawing that same object here guarantees
## the two maps carry identical label text in identical positions -- a spatial
## join back onto the SSURGO polygons would not, because SSURGO's polygon
## boundaries differ from ours.
## ---------------------------------------------------------------------------
if (!exists("series_lab")) {
  stop("Run the soil series chunk first -- it creates `series_lab`.")
}
ssurgo_lab <- series_lab

## ---------------------------------------------------------------------------
## Two distinguishable line networks:
##   dashed grey  = SSURGO map unit boundaries (dissolved on MUKEY)
##   solid black  = our mapped soil series boundaries (the `series` layer, so
##                  they are identical to the soil series map)
## The agreement fill itself is drawn with NO border, otherwise the
## intersection slivers would add a third set of lines.
## ---------------------------------------------------------------------------
ssurgo_units <- dissolve_by(ssurgo, "MUKEY")

ssurgo_map <- tm_shape(ssurgo, bbox = sf::st_bbox(border)) +
  tm_polys_safe(MATCH_FIELD, match_pal, "SSURGO match", legend_ncol = 1,
                border_col = NA, border_lwd = 0) +
  tm_shape(ssurgo_units) +
  tm_borders(col = "grey15", lwd = 1.0, lty = "dashed") +
  tm_shape(series) +
  tm_borders(col = "black", lwd = 0.7, lty = "solid") +
  tm_shape(ssurgo_lab) +
  tm_text_safe("musym", size = 0.38) +
  tm_shape(border) +
  tm_borders(col = "black", lwd = 1.8) +
  tm_line_legend_safe(
    labels = c("SSURGO map unit", "Mapped soil series"),
    col    = c("grey15", "black"),
    lty    = c("dashed", "solid"),
    lwd    = c(1.0, 0.7),
    title  = "Boundaries"
  ) +
  tm_compass_safe(size = 2.6) +
  tm_scalebar_safe() +
  panel_layout("Field Series vs. SSURGO Map Unit -- Agreement")

save_one(ssurgo_map, "Perry_ssurgo_matchup.pdf")             # -> outputs/maps
ssurgo_map
```

```{r ssurgo-table}
#| echo: false

tab <- as.data.frame(table(ssurgo[[MATCH_FIELD]]))
names(tab) <- c("Match status", "Polygons")
tab$Percent <- sprintf("%.1f%%", 100 * tab$Polygons / sum(tab$Polygons))
knitr::kable(tab, caption = "SSURGO agreement summary")
```

\newpage

# Appendix: Layer Inventory

```{r inventory}
#| echo: false

inv <- do.call(rbind, lapply(c(chem_layers, texture_layers), function(s) {
  data.frame(
    Map    = s$title,
    Source = s$file,
    Band   = if (is.null(s$band)) 1L else as.integer(s$band),
    Factor = if (is.null(s$mult)) 1 else s$mult,
    stringsAsFactors = FALSE
  )
}))
inv <- rbind(
  inv,
  data.frame(Map = "Soil series",   Source = "Soil_Series_Map.shp", Band = NA_integer_, Factor = NA_real_),
  data.frame(Map = "SSURGO matchup", Source = "SSURGO_matchup.shp",  Band = NA_integer_, Factor = NA_real_),
  data.frame(Map = "Site border",    Source = "border_17N.shp",      Band = NA_integer_, Factor = NA_real_)
)
knitr::kable(inv, caption = "Layers rendered in this document")
```

All maps are projected to **NAD83 / UTM Zone 17N (EPSG:26917)**, the native CRS
of the kriged surfaces. Scale bars are in metres.

`Factor` is the multiplier applied on read. Ca, K, Mg, Mn, P and Zn are stored
as Mehlich-1 ppm despite the `_lbsac` filename, and are converted to lbs/ac
here using the acre-furrow-slice factor of 2. pH and lime buffer capacity are
not converted.



```{r Package maps}

## ---------------------------------------------------------------------------
## Bundles every layer into one zip: GeoTIFF rasters + ESRI shapefiles.
## Paste anywhere AFTER the `paths` chunk. Re-reads from disk, so chunk order
## does not matter.   install.packages("zip")  # pure R, no Rtools needed
## ---------------------------------------------------------------------------
library(terra); library(sf); library(zip)
sf::sf_use_s2(FALSE)

INCLUDE_VARIANCE <- TRUE      # also ship the kriging variance surfaces
TARGET_CRS       <- 26917     # NAD83 / UTM 17N, native CRS of the kriged grids

if (!exists("dir_krige"))   dir_krige   <- "C:/git/spatial_SCP_taxonomy/outputs/perry_ordinary_kriging_no_covars/clipped_and_lbsac"
if (!exists("dir_series"))  dir_series  <- "C:/git/coastal_plain_taxonomy/outputs/spatial"
if (!exists("path_border")) path_border <- "C:/Users/Madelyn Willis/OneDrive/Documents/perry_git/perry/Mapping/Data/Border/border_17N.shp"
if (!exists("dir_out"))     dir_out     <- "C:/git/spatial_SCP_taxonomy/outputs/maps"
if (!exists("CONVERT_PPM_TO_LBSAC")) CONVERT_PPM_TO_LBSAC <- TRUE

PPM2      <- if (CONVERT_PPM_TO_LBSAC) 2 else 1
chem_unit <- if (CONVERT_PPM_TO_LBSAC) "lbs/ac" else "ppm"

pkg_name <- paste0("UGA_GrandFarm_Perry_soil_layers_", format(Sys.Date(), "%Y%m%d"))
pkg_dir  <- file.path(dir_out, pkg_name)
unlink(pkg_dir, recursive = TRUE)
dir.create(file.path(pkg_dir, "raster"),    recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(pkg_dir, "shapefile"), recursive = TRUE, showWarnings = FALSE)

## --- raster registry -------------------------------------------------------
pkg_rasters <- list(
  list(out="soil_Ca",  file="chem_Ca_OK_lbsac.tif", mult=PPM2, unit=chem_unit, desc="Mehlich-1 extractable calcium"),
  list(out="soil_pH",  file="chem_pH.2_OK.tif",     mult=1,    unit="pH units", desc="Soil pH, 1:1 soil:water"),
  list(out="soil_LBC", file="chem_LBC_1_OK.tif",    mult=1,    unit="index",   desc="Lime buffer capacity"),
  list(out="soil_K",   file="chem_K_OK_lbsac.tif",  mult=PPM2, unit=chem_unit, desc="Mehlich-1 extractable potassium"),
  list(out="soil_Mg",  file="chem_Mg_OK_lbsac.tif", mult=PPM2, unit=chem_unit, desc="Mehlich-1 extractable magnesium"),
  list(out="soil_Mn",  file="chem_Mn_OK_lbsac.tif", mult=PPM2, unit=chem_unit, desc="Mehlich-1 extractable manganese"),
  list(out="soil_P",   file="chem_P_OK_lbsac.tif",  mult=PPM2, unit=chem_unit, desc="Mehlich-1 extractable phosphorus"),
  list(out="soil_Zn",  file="chem_Zn_OK_lbsac.tif", mult=PPM2, unit=chem_unit, desc="Mehlich-1 extractable zinc"),
  list(out="texture_sand", file="texture_OK_normalized_0_100_sum100.tif", band=1, mult=1, unit="percent", desc="Sand fraction, normalised to sum 100"),
  list(out="texture_silt", file="texture_OK_normalized_0_100_sum100.tif", band=2, mult=1, unit="percent", desc="Silt fraction, normalised to sum 100"),
  list(out="texture_clay", file="texture_OK_normalized_0_100_sum100.tif", band=3, mult=1, unit="percent", desc="Clay fraction, normalised to sum 100")
)

var_files <- c(soil_Ca="chem_Ca_OKvar.tif", soil_pH="chem_pH.2_OKvar.tif",
               soil_LBC="chem_LBC_1_OKvar.tif", soil_K="chem_K_OKvar.tif",
               soil_Mg="chem_Mg_OKvar.tif", soil_Mn="chem_Mn_OKvar.tif",
               soil_P="chem_P_OKvar.tif", soil_Zn="chem_Zn_OKvar.tif",
               texture_sand="texture_lab_sand_OKvar.tif",
               texture_silt="texture_lab_silt_OKvar.tif",
               texture_clay="texture_lab_clay_OKvar.tif")

meta <- list()
gtiff <- c("COMPRESS=LZW", "TILED=YES", "BIGTIFF=IF_SAFER")

for (s in pkg_rasters) {
  src <- file.path(dir_krige, s$file)
  if (!file.exists(src)) { message("skipped (missing): ", s$file); next }
  r <- terra::rast(src)
  if (!is.null(s$band)) r <- r[[s$band]]
  if (s$mult != 1) r <- r * s$mult
  names(r) <- s$out
  terra::writeRaster(r, file.path(pkg_dir, "raster", paste0(s$out, ".tif")),
                     overwrite = TRUE, gdal = gtiff, datatype = "FLT4S")
  v <- terra::values(r, mat = FALSE)
  meta[[length(meta)+1]] <- data.frame(
    layer=s$out, type="raster", file=paste0("raster/", s$out, ".tif"),
    units=s$unit, description=s$desc,
    min=round(min(v,na.rm=TRUE),3), max=round(max(v,na.rm=TRUE),3),
    mean=round(mean(v,na.rm=TRUE),3), source=s$file, stringsAsFactors=FALSE)
  message("raster  ", s$out)

  if (INCLUDE_VARIANCE && s$out %in% names(var_files)) {
    vsrc <- file.path(dir_krige, var_files[[s$out]])
    if (file.exists(vsrc)) {
      rv <- terra::rast(vsrc); names(rv) <- paste0(s$out, "_var")
      terra::writeRaster(rv, file.path(pkg_dir, "raster", paste0(s$out, "_variance.tif")),
                         overwrite = TRUE, gdal = gtiff, datatype = "FLT4S")
      meta[[length(meta)+1]] <- data.frame(
        layer=paste0(s$out,"_variance"), type="raster",
        file=paste0("raster/", s$out, "_variance.tif"),
        units="squared prediction units",
        description="Ordinary kriging variance (NOT unit-converted)",
        min=NA_real_, max=NA_real_, mean=NA_real_,
        source=var_files[[s$out]], stringsAsFactors=FALSE)
    }
  }
}

## --- shapefiles ------------------------------------------------------------
## Field names are capped at 10 characters and GDAL truncates silently, so all
## names below are already <= 10 characters.
write_shp <- function(x, name, desc) {
  suppressWarnings(
    sf::st_write(x, file.path(pkg_dir, "shapefile", paste0(name, ".shp")),
                 quiet = TRUE, delete_dsn = TRUE)
  )
  meta[[length(meta)+1]] <<- data.frame(
    layer=name, type="vector", file=paste0("shapefile/", name, ".shp"),
    units=NA_character_, description=desc,
    min=NA_real_, max=NA_real_, mean=NA_real_,
    source=NA_character_, stringsAsFactors=FALSE)
  message("vector  ", name, "  (", nrow(x), " features)")
}

bd <- sf::st_read(path_border, quiet = TRUE)
bd <- sf::st_zm(sf::st_transform(bd, TARGET_CRS), drop = TRUE)
write_shp(sf::st_sf(site = "UGA GrandFarm Perry", geometry = sf::st_geometry(bd)),
          "site_border", "Field boundary of the research site")

se <- sf::st_make_valid(sf::st_zm(sf::st_read(file.path(dir_series, "Soil_Series_Map.shp"), quiet = TRUE), drop = TRUE))
se <- sf::st_transform(se, TARGET_CRS)
sp <- trimws(as.character(se$Series_pri)); sp[is.na(sp) | sp == ""] <- "HTM"
write_shp(sf::st_sf(series = sp,
                    musym  = trimws(as.character(se$musym)),
                    muname = trimws(as.character(se$muname)),
                    geometry = sf::st_geometry(se)),
          "soil_series", "Soil series mapped in the field, with map unit symbol")

ss <- sf::st_make_valid(sf::st_zm(sf::st_read(file.path(dir_series, "SSURGO_matchup.shp"), quiet = TRUE), drop = TRUE))
ss <- sf::st_transform(ss, TARGET_CRS)
raw <- toupper(trimws(as.character(ss$match_stat)))
ss_out <- sf::st_sf(
  match      = ifelse(!is.na(raw) & raw == "AGREE", "Agree", "Disagree"),
  ssgo_musym = trimws(as.character(ss$MUSYM)),
  ssgo_unit  = trimws(as.character(ss$MapUnit)),
  ssgo_ser   = trimws(as.character(ss$Series_cle)),
  fld_series = trimws(as.character(ss$soilseries)),
  drainage   = trimws(as.character(ss$DrainageCl)),
  slope      = trimws(as.character(ss$Slope)),
  geometry   = sf::st_geometry(ss))
write_shp(ss_out, "ssurgo_matchup",
          "Field series vs SSURGO map unit; Disagree includes no-overlap polygons")

keys <- ss_out$ssgo_musym; uk <- unique(keys)
write_shp(sf::st_sf(musym = uk,
                    geometry = do.call(c, lapply(uk, function(k) sf::st_union(ss_out[keys == k, ]))),
                    crs = sf::st_crs(ss_out)),
          "ssurgo_units", "SSURGO map unit polygons, dissolved on map unit symbol")

## --- symbology + metadata --------------------------------------------------
series_cols <- c(Benevolence="#EEDABA", Blanton="#B68870", Dothan="#E6E600",
  Faceville="#FF5500", Greenville="#E64C00", Johns="#ED7932", Lakeland="#FFEBAF",
  Leefield="#F5B05C", Lucy="#F57A7A", Marvyn="#BA7049", Norfolk="#FFAA00",
  Orangeburg="#E69800", `Red Bay`="#A83800", Thursa="#F79965", Troup="#A87000",
  Wagram="#E8D08A", Disturbed="#886868", HTM="#D9D9D9")
rgbm <- grDevices::col2rgb(series_cols)
utils::write.csv(data.frame(series=names(series_cols), hex=unname(series_cols),
                            r=rgbm[1,], g=rgbm[2,], b=rgbm[3,]),
                 file.path(pkg_dir, "soil_series_colors.csv"), row.names = FALSE)
utils::write.csv(do.call(rbind, meta),
                 file.path(pkg_dir, "layer_metadata.csv"), row.names = FALSE)

## --- README ----------------------------------------------------------------
writeLines(c(
"UGA GRANDFARM - PERRY, GEORGIA",
"Soil taxonomic and chemical map layers",
paste0("Packaged ", format(Sys.Date(), "%d %B %Y")),
"",
"COORDINATE REFERENCE SYSTEM -------------------------------------------",
"  All layers: NAD83 / UTM zone 17N (EPSG:26917), metres.",
"  Rasters are 3 m cell size and share an identical grid.",
"",
"CONTENTS ---------------------------------------------------------------",
"  raster/                 GeoTIFF, LZW compressed, Float32, NoData = NaN",
"  shapefile/              ESRI shapefiles (.shp/.dbf/.shx/.prj/.cpg)",
"  layer_metadata.csv      One row per layer: units, description, range",
"  soil_series_colors.csv  Hex/RGB per series, to match the project figures",
"",
"UNITS - PLEASE READ ----------------------------------------------------",
if (CONVERT_PPM_TO_LBSAC)
paste0("  Ca, K, Mg, Mn, P and Zn are in lbs/acre, converted from Mehlich-1\n",
       "  ppm using the acre-furrow-slice factor of 2 (~2,000,000 lb soil per\n",
       "  acre at 6-2/3 in depth). Rescale if your sampling depth differs.")
else
"  Ca, K, Mg, Mn, P and Zn are Mehlich-1 ppm (unconverted).",
"  pH is unitless and is NOT converted.",
"  Lime buffer capacity is as measured and is NOT converted.",
"  Texture layers are percent, normalised so sand+silt+clay = 100.",
"",
"METHOD -----------------------------------------------------------------",
"  Continuous surfaces are ordinary kriging predictions from field sample",
"  points, no covariates, clipped to the site boundary.",
"  *_variance.tif is the kriging variance for that surface, in SQUARED",
"  prediction units and NOT unit-converted, so it is not directly",
"  comparable with the lbs/acre surfaces.",
"",
"  Interpolated surfaces carry uncertainty that grows with distance from",
"  sample points. Check the variance layers before drawing conclusions",
"  about any specific location.",
"",
"OPENING THE DATA -------------------------------------------------------",
"  ArcGIS Pro : Add Data, or drag the .tif / .shp in",
"  QGIS       : drag onto the canvas",
"  R          : terra::rast() and sf::st_read()",
"  Python     : rasterio.open() and geopandas.read_file()",
"",
"  Shapefile field names are capped at 10 characters. See",
"  layer_metadata.csv for what each abbreviated field means.",
"",
"CONTACT ----------------------------------------------------------------",
"  Madelyn Willis - Madelyn.Willis@uga.edu",
"  https://madelynwillis3.github.io/research/"
), file.path(pkg_dir, "README.txt"))

## --- zip -------------------------------------------------------------------
zip_path <- file.path(dir_out, paste0(pkg_name, ".zip"))
if (file.exists(zip_path)) file.remove(zip_path)
zip::zip(zipfile = zip_path,
         files   = list.files(pkg_dir, recursive = TRUE),
         root    = pkg_dir)

cat("\nPackage written:\n  ", zip_path, "\n  ",
    round(file.size(zip_path)/1024/1024, 2), " MB, ",
    length(list.files(pkg_dir, recursive = TRUE)), " files\n", sep = "")
```
```

